High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data\vspace*{\fill}

### 2015-11-24 rmarkdown/prelims_paper.Rnw
made some minor corrections in the paper, and fixed the figures to be named as reps instead of the bam files

@@

While the number of produced ChIP-exo data keeps increasing,
characteristics of ChIP-exo data and optimal strategies for
experimental design and analysis of ChIP-exo data are not fully
investigated yet, including issues of sequence biases inherent to
ChIP-exo data, choice of optimal statistical methods, and
determination of optimal sequencing depth. However, currently the
number of available ChIP-exo data is still limited and their
sequencing depths are still insufficient for such investigation. To
address this limitation we gathered ChIP-exo data from diverse
organisms: CTCF factor in human \cite{exo1}; ER factor in human and
FoxA1 factor in mouse (Serandour et al., 2013 \cite{exoillumina}); and
generated $\sig$ factor in Escherichia coli (E. Coli) under
aerobic ($ + O_2$) condition, and treated by rifampicin by 0 and 20
minutes.

@@

\begin{table}[h]
  \centering
  \caption{Usual quality control indicators applied to the $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient, SSD for standardized standard deviation, NSC for normalized strand cross-correlation coefficient (Another measure related to the strand cross-correlation curve is the RSC was ommited since input sample usually don't exists for ChIP-exo). $\sig$ samples were given by Robert Landick, FoxA1 and ER samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF sample is from Rhee and Pugh, 2011 \cite{exo1}.}

@@

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .9\textwidth]{../figs/for_paper/coverage_diagram.pdf}
  \caption{Diagram of the ChIP-exo Quality control pipeline [this is a
    place-holder]. The genome is partitioned into several regions by
    removing the empty segments and for each regions we calculate a
    collection of QC indicators.}
  \label{fig:qcdiagram}
\end{figure}

@@

We considered the creation of a novel QC measure for ChIP-exo called
local normalized SCC or $\mbox{local-NSC}$.

@@

In ChIP-exo experiments, the background is often digested by the
exonuclease enzyme, therefore to determine the sample's quality is
necessary to address the balance between the enrichment and library
complexity. To diagnose this, we considered the Average Read Coverage
(ARC) and the Unique Read Coverage Rate (URCR) which are defined as:

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate and C)
    Boxplots of the a set of the first replicate regions, stratified
    as in (B) with the $\mbox{local-NSC}$ calculated with the
    different biological replicates.}
  \label{fig:enrich}
\end{figure}

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least \Sexpr{de}
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for the third replicate is \Sexpr{pval[[1]]}, for
first replicate is \Sexpr{pval[[2]]} and for the second is
\Sexpr{pval[[3]]}. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the \Sexpr{de}
level.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth]{../figs/for_paper/sensitivity_exo_olda_data.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/resolution_by_dataset_old_data.pdf}
   \includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/sigma_delta_old_densities.pdf}
   \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/sigma_delta_old_densities.pdf}

@@

\subsection{GC content and mappability quality indicators}
\label{sec:gc_map_qc}

@@

\subsection{Quality control for ChIP-Nexus}
\label{sec:nexus}

@@

\subsection{Analysis of E. Coli's Transcription Initiation Complexes}
\label{sec:ecoli}

@@

\bibliographystyle{plain} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{chip_exo_paper}
\nocite{exo_gb}
\nocite{maplot1}
\nocite{maplot2}

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

@@

While the number of produced ChIP-exo data keeps increasing,
characteristics of ChIP-exo data and optimal strategies for
experimental design and analysis of ChIP-exo data are not fully
investigated yet, including issues of sequence biases inherent to
ChIP-exo data, choice of optimal statistical methods, and
determination of optimal sequencing depth. However, currently the
number of available ChIP-exo data is still limited and their
sequencing depths are still insufficient for such investigation. To
address this limitation we gathered ChIP-exo data from diverse
organisms: CTCF factor in human \cite{exo1}; ER factor in human and
FoxA1 factor in mouse (Serandour et al., 2013 \cite{exoillumina}); and
generated $\sig$ factor in Escherichia Coli (E. Coli) under
aerobic ($ + O_2$) condition, and treated by rifampicin by 0 and 20
minutes.

@@

@@

\begin{table}[H]
  \centering
  \caption{Usual quality control indicators applied to the $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient, SSD for standardized standard deviation, NSC for normalized strand cross-correlation coefficient (Another measure related to the strand cross-correlation curve is the RSC was ommited since input sample usually don't exists for ChIP-exo). $\sig$ samples were given by Robert Landick, FoxA1 and ER samples are from Serandour et al., 2013 \cite{exoillumina} and CTCF sample is from Rhee and Pugh, 2011 \cite{exo1}.}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/coverage_diagram2.pdf}
  \caption{Diagram of the ChIP-exo Quality control pipeline. The
    genome is partitioned into several regions by removing the empty
    segments and for each regions we calculate a collection of QC
    indicators, then for each region $R_k$ a vector of summary
    statistics $T(R_k)$ is calculated.}
  \label{fig:qcdiagram}
\end{figure}

@@

We considered the creation of a novel QC measure for ChIP-exo called
local normalized SCC or $\mbox{local-NSC}$. This measure

\begin{align}
  y(\delta) = r\left[ n^+ \left(x + \frac{\delta}{2}
    \right), n^- \left( x- \frac{\delta}{2} \right)\right]
\nonumber
\end{align}

where again $r$ is the Pearson correlation but $n^+$ and $n^-$ are the
coverage vectors of a given region in the genome. Figure
\ref{fig:locSCC} shows that as the library complexity of the region is
lower, then the local SCC seems to show higher variance. Hence, we
fit a nonparametric regression model to smooth the SCC signal:

\begin{align}
 y(\delta) &= f(x_\delta) + \epsilon_\delta \nonumber
\end{align}

where $y(\delta)$ is the observed SCC for shift $x_\delta$,
$\epsilon_\delta \sim N(0,\sigma_f^2)$. Finally the $\mbox{local-NSC}$
is defined as a signal-to-noise ratio by using:

\begin{align}
    \mbox{local-NSC} &= \frac{max_{x_\delta} \hat{f}(x_\delta)}{\hat{\sigma}_f} \label{locNSC}
\end{align}

\begin{figure}[H]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/local_SCC_example.pdf}
  \caption{Using the FoxA1 sample from \cite{exoillumina}. A) shows an
    example region for the 3 replicates. The library complexity seem
    to be comparable between the two top rows and both outperfom the
    last row. The number of reads for the first replicate is higher
    than the number of reads for the other replicates. B) Shows the
    local SCC for the 3 replicates, this measure is comparable among
    the 3 replicates. The first replicate shows the lowest residual
    error, while the third one shows the highest. In blue, a ``loess''
    regression model is fitted.}
\label{fig:locSCC}
\end{figure}

@@

In ChIP-exo experiments, the background is often digested by the
exonuclease enzyme, therefore to determine the sample's quality is
necessary to address the balance between the enrichment and library
complexity. To diagnose this, we consider the Average Read Coverage
(ARC) and the Unique Read Coverage Rate (URCR) which are defined as:

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \includegraphics[width = .7\textwidth,page =3]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata.pdf}
%  \includegraphics[width = .7\textwidth,page =4]{../figs/Carroll_mice_for_paper/Local_SCC_indicator_by_strata_sameRegion.pdf}
  \caption{Using the mouse-FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, in general
    we can see a slight separation into two strong arms, one
    corresponds to low $\mbox{ARC}$ and varying $\mbox{URCR}$, and for
    the other $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases B)
    Boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    experiments divided by condition and biological replicate.}
  \label{fig:enrich}
\end{figure}

% and C) Boxplots of the a set of the first replicate regions,
% stratified as in (B) with the $\mbox{local-NSC}$ calculated with the
% different biological replicates.

@@

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least \Sexpr{de}
fragments into into the ones that overlap with any of the sample's
peaks. Then we considered a Kolmogorov-Smirnov tests to assess if the
distribution of both groups were the same. The p.value for the third
replicate's test is \Sexpr{pval[[1]]}, for the first replicate's test
is \Sexpr{pval[[2]]} and for the second test is
\Sexpr{pval[[3]]}. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the \Sexpr{de}
level.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth]{../figs/for_paper/sensitivity_exo_olda_data.pdf}
  \includegraphics[width = .46\textwidth]{../figs/for_paper/resolution_by_dataset_old_data.pdf}
\end{figure}

\begin{figure}[H]
\centering
   \includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/sigma_delta_old_densities.pdf}
   \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/sigma_delta_old_densities.pdf}

@@

\subsection{Some additional thoughts for this analysis}

\subsubsection{GC content and mappability quality indicators}
\label{sec:gc_map_qc}

@@

\subsubsection{Quality control for ChIP-Nexus}
\label{sec:nexus}

@@

\subsection{Analysis of E. Coli Transcription Initiation Complexes}
\label{sec:ecoli}

The data showed in figure \ref{fig:exo_example} is actually part of a
more interesting experiment. We have 2 biological replicates of the
$\sig$, $\beta'_f$ and $\beta$ transcription factors under two
conditions, one where rifampicin was applied by 20 minutes and another
where it wasn't applied. As seen in the figure, there is ChIP-exo data
for the 3 factors and PET ChIP-Seq for $\sig$ and $\beta'_f$ (in the
figure both replicates were pooled).

$\sig$ factor is a transcription initiation factor of housekeeping
genes in E. Coli, and both $\beta$ factors transcribe the DNA into
mRNA. Hence, we want to use this data to have a better understanding
of the transcription in E. Coli.

The steps in formation of productive transcription complexes are:

\begin{enumerate}
\item Binding of RNAP to rpomoter DNA to form a closed complex.
\item Rearrangement of RNAP-DNA contacts to form an open complex.
\item Binding of nucleoside triphosphates in the active site and
  synthesis of a initial RNA to yield an initial transcription complex
  (ITC) - this is sometimes called a ``scrunched complex''.
\item Promoter escape to form a elongated complex (EC$_1$) when the
  promoter contacts are broken, and the $\sig$ factor is realeased.
\item Conversion of EC$_1$ into EC$_2$ by loading of the ribosome and
  elongation factors to enable efficient transcript elongation.
\end{enumerate}

% \begin{figure}[H]
% \centering
% \begin{tikzpicture}[node distance = 3cm]
% \tikzstyle{region} = [draw,thin,fill = blue!20]
% \tikzstyle{line} = [draw,thick]
% %\path[use as bounding box] (-1,0) rectangle (10,-2);
% \path[-] node[region] (step1) {RNAP + Promoters};
% \path[-] node[region,right of =step1,fill = red!20] (step2) {Closed};
% \path[-] node[region,right of =step2,fill = green!20] (step3) {Intermediate};
% \path[-] node[region,right of =step3,fill = yellow!20] (step4) {Open};
% \path[line][<->] (step1) -- (step2);
% \path[line][<->] (step2) -- (step3);
% \path[line][<->] (step3) -- (step4);
% \end{tikzpicture}
% \caption{Zooming into transcription}
% \end{figure}

As a proof of concept we were able to identify open and closed regions
by centering the binding the peaks respecto to the highest $\sig$
summit (among the rif0 and rif20 samples), and then using hierachical
clustering on the $\beta'_f$ signal of the rif0 samples to produce:

\begin{figure}[H]
  \centering
 \includegraphics[width = \textwidth]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/filtered/heatmaps_centroid.png}
 \caption{Centered and normalized heatmaps for PET ChIP-Seq peaks of
   $\sig$ and $\beta'_f$ under rif0 and rif20. The regions are
   organized by the centroids clustering of the $\beta'_f$-rif0
   profiles.}
  \label{fig:hm1}
\end{figure}

Figure \ref{fig:hm1} shows heatmaps of the normalized coverages of PET
ChIP-Seq peaks for $\beta'_f$ and $\sig70$ under rif0 and rif20
treatments. Each row represents the same region, under the four
different samples. As expected the signal for both $\sig$ samples is
centered around the middle, for $\beta'_f$-rif20, the majority of he
signal seems to be centered, hence the trascription was interrupted
and for $\beta'_f$-rif0 we can see clusters formed by open regions.

Then, using a set of high quality PET ChIP-Seq peaks and clustering
with the mclust model by Fraley and Raftery, 2007 \cite{mclust} we
were able to label the a set of regions as closed or open
complexes. Hence, we expect that by using ChIP-exo data we may find
different patterns that represent transcription units whose output was
limited at differemt steps in the formation of EC$_2$.

Additionally if we find several cases of those patterns, then we may
be able to determine which sequences may be causing this behaviour at
the different steps.

\subsection{Enhancer prediction and active learning}
\label{sec:enhancer}

Last summer, we entered to a contest from ENCODE's functional
characterization group to fit a model to classify enhancers. Roughly
speaks, the data set consided of a set of regions
$\mathcal{R}_1,\cdots,\mathcal{R}_n$ and a set of labels
$Y_1,\cdots,Y_n \in \{0,1,\}$. Hence, we built a series of features
$X_1,\cdots,X_p$ and fitted a model by dividing the $n$ observations
into a training set and a test set, etc. However, there are in total
another $m >> n$ regions in the experiment, and the cost of labelling
those $m$ regions may be high (in either the actual monetary cost of
the experiment or the time the scientists spends labelling those
regions). Those $n+m$ regions are H3K27ac peaks, hence they were
ranked and the first $n$ were selected randomly from groups with high,
medium and low signal. This raises the following questions:

\begin{itemize}
\item Is there a better way of ranking this $n+m$ regions ?
\item What is the best way to select the regions that are going to be
  labelled, i.e. we would want to label the regions that gives us the
  best model.
\end{itemize}

Hence, it may be worth to investigate which active learning strategies
may be the best to apply in this particular problem. Settles, 2012
\cite{al} is a survey of acive learning techniques that we want to
explore in order to solve this problems.

@@

\bibliographystyle{plain} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{chip_exo_paper}
\nocite{exo_gb}
\nocite{maplot1}
\nocite{maplot2}
\nocite{esl}
\nocite{anp}
\nocite{al}
\nocite{oc_review}
\nocite{exo4}

### 2015-11-24 rmarkdown/prelims_paper.tex
made some minor corrections in the paper, and fixed the figures to be named as reps instead of the bam files

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. The p.value for the third replicate is 0.1039, for
first replicate is 0.0064 and for the second is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the 100
level.

>>>>

In order to assess the strand imbalance we created the following
visualization presented in figure \ref{fig:strand}. A) shows the FSR's
behavior as lower depth regions are being filtered out, while B) shows
which percentage of the regions are composed by fragments of both
strands or only one (forward or backward). As the number of reads per
region's lower bound increases, the quantiles tend to approach the
median. To evaluate this statement, for each replicate of the FoxA1
dataset, we divided the regions formed by at least 100
fragments into into the ones that overlap with any of the sample's
peaks. Then we considered a Kolmogorov-Smirnov tests to assess if the
distribution of both groups were the same. The p.value for the third
replicate's test is 0.1039, for the first replicate's test
is 0.0064 and for the second test is
0.0871. This shows that only the middle sample from figure
\ref{fig:strand} is the least imbalance sample from the the three,
while the other two still show a higher imbalance at the 100
level.
